Linear Vlasov Theory in the Shearing Sheet Approximation \\ with Application to the Magneto-Rotational Instability

### 2014-08-14 paper.tex
Added missing factor of ωc² in eq. (73)
Made proof corrections

\RequirePackage[l2tabu, orthodox]{nag}
\documentclass[aps,pre,notitlepage,amsmath,amssymb,amsfonts,nobibnotes,nofootinbib]{revtex4-1}
\setcitestyle{authoryear,round}
\bibliographystyle{apsrev4-1}

@@

\author{Tobias Heinemann}
\affiliation{University of California at Berkeley}
\affiliation{Kavli Institute for Theoretical Physics}
\author{Eliot Quataert}
\affiliation{University of California at Berkeley}

@@

This paper is organized as follows. We begin by providing a brief overview of
kinetic theory in the  shearing sheet approximation (\cref{sec:vlasov-sheet}).
We then describe in some detail the equilibrium solution of the shearing sheet
equations, emphasizing in particular the properties of the equilibrium
distribution function (\cref{sec:steady-state}).
\Cref{sec:conductivity-tensor} contains the key derivation of the conductivity
tensor and the linear dispersion relation in the shearing sheet and shows how
those can be derived from standard uniform plasma results in the literature by
an appropriate set of coordinate transformations.
\Cref{sec:cold,sec:MRI-finite-cyc} derive analytically and determine
numerically the MRI dispersion relation in various limits and discuss the
connection between our results and previous derivations in the literature.
Finally, \cref{sec:conclusions} summarizes our key results. The Appendices
contain various technical aspects of the calculations.

@@

In the absence of Coriolis and tidal forces, $\Lambda_{+}$ ($\Lambda_{-}$)
would describe a right (left) handed circularly polarized wave. Insertion of
\cref{eq:parallel-response} with \cref{eq:circular-components} into
\cref{eq:perpendicular-dispersion-single} yields the perpendicular dispersion
tensor
\begin{equation}
  \label{eq:D-perp}
  \mathbf{D}_\perp \doteq
  k_z^2\va^2\begin{pmatrix} 1 & 0 \\ 0 & 1\end{pmatrix}
  + \begin{pmatrix}
    1 & 0 \\ 0 & \omg/S_z
  \end{pmatrix}
  \cdot\begin{pmatrix}
    \Re & \Im \\ -\Im & \Re
  \end{pmatrix}
  \cdot\begin{pmatrix}
    1 & 0 \\ 0 & \omg/S_z
  \end{pmatrix}
  - i\omega\omc b_z\begin{pmatrix} 0 & -1 \\ 1 & 0\end{pmatrix}
  -2q\Omega^2\frac{\omc b_z}{S_z}\begin{pmatrix} 0 & 0 \\ 0 & 1\end{pmatrix},
\end{equation}
where the right hand side is the matrix representation of $\mathbf{D}_\perp$
in the basis $(\ex,\ey)$. We have also introduced the short hands
\begin{equation}
  \Re = \frac{1}{2}(\Lambda_{+} + \Lambda_{-})
  \quad\textrm{and}\quad
  \Im = \frac{1}{2i}(\Lambda_{+} - \Lambda_{-}).
\end{equation}
We note that $\Lambda_{-}=\Lambda_{+}^\ast$ for purely imaginary $\omega$. In
this case, $\Re$ and $\Im$ are simply the real and imaginary parts of
$\Lambda_{+}$.

@@

The integrals of motion are the particle energy
\begin{equation}
  \mathcal{E} = \frac{\dot{r}^2 + r^2\dot{\varphi}^2 + \dot{z}^2}{2}
  + \Psi + \cm\Phi
\end{equation}
and the canonical angular momentum $p_\varphi$ defined in
\cref{eq:canonical-angular-momentum}. Let us define the guiding center radius
$\rc$ through
\begin{equation}
  p_\varphi = \rc^2\Omega(\rc,0) - \cm\psi(\rc,0).
\end{equation}
Thus, $\rc$ is the radius of circular orbit with angular momentum $p_\varphi$.
Its energy is
\begin{equation}
  \mathcal{E}_c = \frac{\rc^2\Omega^2(\rc,0)}{2}
  + \Psi(\rc,0) + \cm\Phi(\rc,0).
\end{equation}

Note that $\mathcal{E}_c$ depends only on the angular momentum and is thus an
integral of motion. By Jeans' theorem, any distribution function of the form
$f(\mathcal{E},\mathcal{E})$ is thus a solution to the steady state Vlasov
equation. We define the modified electromagnetic Schwarzschild distribution as
\begin{equation}
  \label{eq:modified-em-schwarzschild}
  f \propto \exp(-\mathcal{K}/T),
\end{equation}
where
\begin{equation}
  \label{eq:global-gyration-energy}
  \mathcal{K} = \mathcal{E} - \mathcal{E}_c
\end{equation}
is the gyration energy. The proportionality constant in
\cref{eq:modified-em-schwarzschild} and the temperature $T$ may in general
depend on the canonical angular momentum.

@@

@@

where we have introduced the tensor

@@

where we have introduced a new time variable $\tau=t-t'$ and the short hand

@@

The electrons dynamics are described by the momentum equation
\begin{equation}
  \label{eq:ohms-law}
  e n_e(\vec{E} + \ue\times\vec{B}) + \nabla\pe = 0,
\end{equation}
which we will henceforth refer to as the generalized Ohm's law. Throughout
this work we assume that the electron pressure $\pe$ is a function of the
electron number density $n_e$ alone, i.e.\ we assume a barotropic equation of
state. \Cref{eq:ohms-law} arises out of the electron momentum equation in the
limit of vanishing electron mass and assuming an isotropic pressure. The
electrons must be collisional for this assumption to be reasonable. Indeed,
\citet{Rosin2011} derive \cref{eq:ohms-law} rigorously through an expansion in
the electron-to-ion mass ratio under the assumption of weakly collisional ions
but strongly collisional electrons.

>>>>

\RequirePackage[l2tabu, orthodox]{nag}
\documentclass[aps,pre,notitlepage,amsmath,amssymb,amsfonts,nobibnotes,nofootinbib,superscriptaddress]{revtex4-1}
\setcitestyle{authoryear,round}
\bibliographystyle{apsrev4-1}

@@

\author{Tobias Heinemann}
\email{tobias.heinemann@gmail.com}
\affiliation{Department of Astronomy,
  University of California, Berkeley, CA 94720, USA}
\affiliation{Kavli Institute for Theoretical Physics,
  University of California, Santa Barbara, CA 93106, USA}
\author{Eliot Quataert}
\email{eliot@berkeley.edu}
\affiliation{Department of Astronomy,
  University of California, Berkeley, CA 94720, USA}

@@

This paper is organized as follows. We begin by providing a brief overview of
kinetic theory in the  shearing sheet approximation (\cref{sec:vlasov-sheet}).
We then describe in some detail the equilibrium solution of the shearing sheet
equations, emphasizing in particular the properties of the equilibrium
distribution function (\cref{sec:steady-state}).
\Cref{sec:conductivity-tensor} contains the key derivation of the conductivity
tensor and the linear dispersion relation in the shearing sheet and shows how
those can be derived from standard uniform plasma results in the literature by
an appropriate set of coordinate transformations.
\Cref{sec:cold,sec:MRI-finite-cyc} derive analytically and determine
numerically the MRI dispersion relation in various limits and discuss the
connection between our results and previous derivations in the literature.
Finally, \cref{sec:conclusions} summarizes our key results. The Appendices
\labelcref{app:unperturbed-hamiltonian} through \labelcref{app:vlasov-fluid}
contain various technical aspects of the calculations.

@@

In the absence of Coriolis and tidal forces, $\Lambda_{+}$ ($\Lambda_{-}$)
would describe a right (left) handed circularly polarized wave. Insertion of
\cref{eq:parallel-response} with \cref{eq:circular-components} into
\cref{eq:perpendicular-dispersion-single} yields the perpendicular dispersion
tensor
\begin{equation}
  \label{eq:D-perp}
  \mathbf{D}_\perp \doteq
  k_z^2\va^2\begin{pmatrix} 1 & 0 \\ 0 & 1\end{pmatrix}
  + \omc^2\begin{pmatrix}
    1 & 0 \\ 0 & \omg/S_z
  \end{pmatrix}
  \cdot\begin{pmatrix}
    \Re & \Im \\ -\Im & \Re
  \end{pmatrix}
  \cdot\begin{pmatrix}
    1 & 0 \\ 0 & \omg/S_z
  \end{pmatrix}
  - i\omega\omc b_z\begin{pmatrix} 0 & -1 \\ 1 & 0\end{pmatrix}
  -2q\Omega^2\frac{\omc b_z}{S_z}\begin{pmatrix} 0 & 0 \\ 0 & 1\end{pmatrix},
\end{equation}
where the right hand side is the matrix representation of $\mathbf{D}_\perp$
in the basis $(\ex,\ey)$. We have also introduced the short hands
\begin{equation}
  \Re = \frac{1}{2}(\Lambda_{+} + \Lambda_{-})
  \quad\textrm{and}\quad
  \Im = \frac{1}{2i}(\Lambda_{+} - \Lambda_{-}).
\end{equation}
We note that $\Lambda_{-}=\Lambda_{+}^\ast$ for purely imaginary $\omega$. In
this case, $\Re$ and $\Im$ are simply the real and imaginary parts of
$\Lambda_{+}$.

@@

The integrals of motion are the particle energy
\begin{equation}
  \mathcal{E} = \frac{\dot{r}^2 + r^2\dot{\varphi}^2 + \dot{z}^2}{2}
  + \Psi + \cm\Phi
\end{equation}
and the canonical angular momentum $p_\varphi$ defined in
\cref{eq:canonical-angular-momentum}.  By Jeans' theorem, any distribution
function of the form $f(\mathcal{E},p_\varphi)$ is a solution of the steady
state Vlasov equation. Let us define the guiding center radius $\rc$ through
\begin{equation}
  p_\varphi = \rc^2\Omega(\rc,0) - \cm\psi(\rc,0).
\end{equation}
Thus, $\rc$ is the radius of circular orbit with angular momentum $p_\varphi$.
Its energy is
\begin{equation}
  \mathcal{E}_c = \frac{\rc^2\Omega^2(\rc,0)}{2}
  + \Psi(\rc,0) + \cm\Phi(\rc,0).
\end{equation}

Note that $\mathcal{E}_c$ depends only on the angular momentum and is thus an
integral of motion. We define the modified electromagnetic Schwarzschild
distribution as
\begin{equation}
  \label{eq:modified-em-schwarzschild}
  f \propto \exp(-\mathcal{K}/T),
\end{equation}
where
\begin{equation}
  \label{eq:global-gyration-energy}
  \mathcal{K} = \mathcal{E} - \mathcal{E}_c
\end{equation}
is the gyration energy. The proportionality constant in
\cref{eq:modified-em-schwarzschild} and the temperature $T$ may in general
depend on the canonical angular momentum.

@@

  \cdot\delta\tvec{E}.

@@

Here, we have introduced the tensor

@@

where the new time variable $\tau=t-t'$ and the short hand

@@

The electrons dynamics are described by the momentum equation
\begin{equation}
  \label{eq:ohms-law}
  e n_e(\vec{E} + \ue\times\vec{B}) + \nabla\pe = 0,
\end{equation}
which we will henceforth refer to as the generalized Ohm's law. Here, we
assume that the electron pressure $\pe$ is a function of the electron number
density $n_e$ alone, i.e.\ we assume a barotropic equation of state.
\Cref{eq:ohms-law} arises out of the electron momentum equation in the limit
of vanishing electron mass and assuming an isotropic pressure. The electrons
must be collisional for this assumption to be reasonable. Indeed,
\citet{Rosin2011} derive \cref{eq:ohms-law} rigorously through an expansion in
the electron-to-ion mass ratio under the assumption of weakly collisional ions
but strongly collisional electrons.
